We have implemented an interpreter and compiler for the RhoStratego
language described in chapter \ref{ch:rhostratego}, as well as an
implementation of the type checker given in chapter
\ref{ch:typesystem}.  

The RhoStratego implementation consists of the following components:
\begin{itemize}
\item A parser for the RhoStratego language, generated from a grammar
  specified using the Syntax Definition Formalism (SDF2)
  \cite{syntaxdefinition}.  The grammar is translated to a parse table
  using the \code{sdf2table} program, which is part of XT.  This table
  can then be used with the \code{sglr} and \code{implode-asfix}
  programs --- also part of XT --- to parse a file into an ATerm.
\item A desugarer that takes the output of the parser and simplifies
  it.
\item A pretty-printer, to transform RhoStratego abstract syntax to
  concrete syntax.
\item An interpreter written in Stratego, implementing both the lazy
  and strict evaluation strategies.  The interpreter is quite limited
  --- it does not implement most I/O functions, for example --- and is
  primarily useful for language experiments.  The source code of the
  interpreter is given in appendix \ref{ap:interpreter}.
\item A type checker written in Stratego.  It reads a desugared
  RhoStratego module in ATerm format, infers (as far as possible) and
  checks types, and produces a desugared RhoStratego module in ATerm
  format in which every expression has been decorated with its type.
\item A compiler written in Stratego.  It translates desugared
  RhoStratego modules in ATerm format to C code.
\item A runtime system and standard library against which RhoStratego
  programs must be linked.
\item A set of test cases (i.e., RhoStratego programs along with their
  expected outputs) to test the compiler against automatically.
\end{itemize}

The source code of the RhoStratego implementation can be obtained
through anonymous CVS.  On Unix systems this can be accomplished using
the following commands:
\begin{verbatim}
% export CVSROOT=\
 :pserver:anoncvs@losser.st-lab.cs.uu.nl:/home/cvsrepo
% cvs login
[Password: anon]
% cvs -z9 checkout rho
\end{verbatim}
A web interface to the CVS repository is available at
\url{http://losser.st-lab.cs.uu.nl/services/cvsweb/rho/}.  Further
information can be obtained at
\url{http://www.stratego-language.org/rho/}.